formula1

### 2020-06-23 inst/f1points.Rmd
more work

## Formula 1 racing: points systems

@@

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on (Hartley,
Button, and Resta placed last).  In the first column, we see the
result from Australia (AUS) in which Hamilton came second, Vettel
first, Bottas third, and so on.  Notation used also includes six
different classes of no-score such as `Ret` for retired, `WD` for
withdrawn, and so on; a zero entry means "did not finish".  It is
straightforward to translate this table into a Plackett-Luce
log-likelihood function using the `hyper2` package; for simplicity we
will consider only the `r top` top-ranked drivers\footnote{In the
Plackett-Luce likelihood function, the performance of lower-ranked
players can be weakly informative about higher-ranked players'
strengths.  For example, we see that Vettel retired twice (in
Singapore and Japan), so any player who placed in those venues will
effectively ``steal'' strength from Vettel, and generally ``give'' it
to Hamilton or Bottas.}.  Although it has many terms, the first few
are

```{r,label=showanordertableandhead,echo=FALSE,print=FALSE}
library(hyper2,quietly=TRUE)
top <- 11
H <- ordertable2supp(F1_table_2017[seq_len(top),])
head(H,5)
```

Finding the maximum likelihood estimate for the players' strengths is
straightforward:

```{r,label=printapiechart,echo=FALSE}
mH <- maxp(H)
print(mH)
pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
```

We see the Hamilton has the largest estimated strength, although it
should be noted that a likelihood ratio test [`samep.test()`, supplied
with the package] does not reject the null of
$p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ , with a likelihood ratio of
$e^{1.5}\simeq 4.48$, corresponding to an asymptotic $p$-value of
about $0.08$.

@@

@@

Have a quick look at the result:

```{r,label=quicklookatOO}
dim(OO)
rownames(OO) <- names(pointslist)
plot(apply(OO,1,mean),ylab='mean number of ranks correct')
```

And more work:

```{r,label=multiplemontecarlotabulation,echo=FALSE}
M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
dimnames(M3) <- list(system=names(pointslist),correct=0:11)
head(M3)
```

now cumulative:

```{r,label=cummulttab,echo=FALSE}
Md <- t(apply(M3,1,cumsum))
colnames(Md) <- paste("<=",colnames(Md),sep="")
Md
```

@@

```{r,label=plotcumulativeM,fig=TRUE}
plot(Md[,1],ylab='number wrong')
plot(1000-Md[,11],ylab='number perfect')
matplot(t(Md),type="b",lty=1,pch=16,col=rainbow(nrow(Md)))
legend("bottomright",lty=1,pch=16,col=rainbow(nrow(Md)),legend=letters[seq_len(nrow(Md))])
plot(points_real)
grid()
```

```{r,makeatable}
table(real_winner_ok=OO[1,]>0,wta_winner_ok=OO[3,]>0)
table(real_all=OO[1,]==11,wta_all=OO[3,]==11)
write.csv(OO,file="~/f.txt")
```

Consider the first column of the table above which shows the number of
virtual races in which each scoring system gets the winner incorrect;
a good points system has a low value here.  Note the peculiar finding
that `top3` has by far the lowest number, a finding which I have no
explanation for.

>>>>

@@

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on (Hartley,
Button, and Resta placed last).  In the first column, we see the
result from Australia (AUS) in which Hamilton came second, Vettel
first, Bottas third, and so on.  Notation used also includes six
different classes of no-score such as `Ret` for retired, `WD` for
withdrawn, and so on; a zero entry means "did not finish".  It is
straightforward to translate this table into a Plackett-Luce
log-likelihood function using the `hyper2` package; for simplicity we
will consider only the `r top` top-ranked drivers\footnote{In the
Plackett-Luce likelihood function, the performance of lower-ranked
players can be weakly informative about higher-ranked players'
strengths.  For example, we see that Vettel retired twice (in
Singapore and Japan), so any player who placed in those venues will
effectively ``steal'' strength from Vettel, and generally ``give'' it
to Hamilton or Bottas.}.  Although it has many terms, the overall loglikelihood 
expression is of the general form

```{r,label=showanordertableandhead,echo=FALSE,print=FALSE}
library(hyper2,quietly=TRUE)
top <- 11
H <- ordertable2supp(F1_table_2017[seq_len(top),])
```

\newcommand{\pham}{p_\mathrm{Ham}}
\newcommand{\pvet}{p_\mathrm{Vet}}
\newcommand{\pbot}{p_\mathrm{Bot}}
\newcommand{\prai}{p_\mathrm{Rai}}
\newcommand{\pric}{p_\mathrm{Ric}}
\newcommand{\pver}{p_\mathrm{Ver}}
\newcommand{\pper}{p_\mathrm{Per}}
\newcommand{\poco}{p_\mathrm{Oco}}
\newcommand{\psai}{p_\mathrm{Sai}}
\newcommand{\phul}{p_\mathrm{Hul}}
\newcommand{\pmas}{p_\mathrm{Mas}}

\begin{equation}
\frac{\pham^{20}\,\pmas^{16}\,\pbot^{19}\ldots
}{\parbox{4in}{$(\pper+\poco)(\pver+\pper+\psai+\phul+\pmas)\\
\rule{10mm}{0mm}(\pric+\pper+\pmas)(\psai+\pmas)\\ \rule{20mm}{0mm}
         (\pver + \pper+ \poco + \psai+ \phul)\ldots
	 $}}
 \end{equation}

Finding the maximum likelihood estimate for the players' strengths is
straightforward.  The `hyper2` package includes a suite of numerical
optimization routines, and because they have access to derivatives,
convergence is rapid.

```{r,label=printapiechart,echo=FALSE}
mH <- maxp(H)
mH
pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
```

We see the Hamilton has the largest estimated strength (at about
0.39), followed by Vettel and then Bottas at about 0.2 and 0.13
respectively.  However, it should be noted that a likelihood ratio
test [`samep.test()`, supplied with the package] does not reject the
null that Hamilton and Vettel have the same strength ($H_0\colon
\pham= \pvet$) with a likelihood ratio of $e^{1.5}\simeq 4.48$,
corresponding (by Wilks's theorem) to an asymptotic $p$-value of about
$0.08$.

@@

(note carefully that the historically correct points awarded to the
drivers differs from that calculated here.  That is for two reasons:
firstly, ``fastest lap'' points are not included here, and also the
truncation of the order table to the first 11 drivers can increase the
rank of a driver if non-first-11 drivers are placed).

@@

now some summaries:

```{r,label=somesummaries}
summarytable <- data.frame(
means = apply(OO,1,mean),
winner_correct = rowSums(OO>1),
all_correct = rowSums(OO==top))
rownames(summarytable) <- names(pointslist)
summarytable
```

```{r,fig=TRUE}
typecol <- c(rep("red",2),rep("orange",10),rep("blue",11),rep("orange",1),rep("black",7),rep("gray",7))
plot(summarytable[,1],ylab='mean',pch=16,col=typecol)
plot(summarytable[,2],ylab='prob of winner correct',pch=16,col=typecol)
plot(summarytable[,3],ylab='prob. of total rank order correct',pch=16,col=typecol)
```

@@

```{r,label=plotcumulativeM,fig=TRUE}
o <- t(apply(1+OO,1,tabulate))
#matplot(0:11,apply(o[,12:1],1,cumsum)/1e3,type='b',lty=1,pch=16,ylab='probability',axes=FALSE,xlab='number incorrect')
jj <- c(0,2,4,6,8,11)
#axis(1,pos=0,at=jj,labels=c("0",paste("<=",jj[-1],sep="")))
#axis(2)
```

```{r,makeatable}
table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
save(OO,file="~/f.Rd")
```

Consider the first column of the table above which shows the number of
virtual races in which each scoring system gets the winner incorrect;
a good points system has a low value here.  Note that `top3` has by
far the lowest number, a finding which I have no explanation for.
